The illusory promise of the Aligned Rank Transform

### 2025-12-18 appendix_I.qmd
methodology..

---
title: "The illusory promise of the Aligned Rank Transform"
subtitle: "Appendix I. Mathematical explanations and proofs"
author: 
  - name: Theophanis Tsandilas
    orcid: 0000-0002-0158-228X
    email: theophanis.tsandilas@inria.fr
    affiliations:
      - name: Université Paris-Saclay, CNRS, Inria, LISN
        country: France
  - name: Géry Casiez
    orcid: 0000-0003-1905-815X
    email: gery.casiezuniv-lille.fr
    affiliations:
      - name: Univ. Lille, CNRS, Inria, Centrale Lille, UMR 9189 CRIStAL
        country: France
bibliography: bibliography.bib
csl: canadian-journal-of-philosophy.csl

@@

## Comparing Type I error rates under non-linear models
Consider again the general relationship between the expected value of a response variable $Y$, which may follow a non-normal distribution, and the predictors in an experimental design with two independent variables $x_1$ and $x_2$:

$$
E(Y|x_1, x_2) = g^{-1}(a_0 + a_1 x_1 + a_2 x_2 + a_{12}x_1 x_2)
$$
where $g^{-1}$ is the inverse link function.

We show that, in the main scenarios considered in this article, the interpretation of effects is not affected by whether the null hypothesis is defined on the coefficient scale or on the response-mean scale. Our analysis follows related discussions in @Ai:2003, @Greene:2010, and @McCabe:2022.

For clarity, let
$$
\eta(x_1, x_2) = a_0 + a_1 x_1 + a_2 x_2 + a_{12}x_1 x_2
$$
denote the linear predictor, and
$$
\mu(x_1, x_2) = E(Y|x_1, x_2) = g^{-1}(\eta(x_1, x_2))
$$
denote the expected value (or mean) as a function of $x_1$ and $x_2$.

### Main effects
We first show that there is no ambiguity in the interpretation of main effects when $a_{12}=0$.

Without loss of generality, assume that $a_1 \neq 0$ an examine the effect on the second factor $x_2$ when $a_2 = 0$. In that case,
$$
\begin{aligned}
\eta(x_1, x_2) = a_0 + a_1 x_1, \qquad
\mu(x_1, x_2) = g^{-1}(a_0 + a_1 x_1)
\end{aligned}
$$
so $\mu$ is a function of $x_1$ only.

We consider two situations:

**(a) Continuous $x_2$: effect defined by a partial derivative**

The effect of $x_2$ on the response can be defined as $\frac{\partial \mu}{\partial x_2}$. Because the argument of $g^{-1}$ contains no $x_2$:
$$\frac{\partial \mu(x_1, x_2)}{\partial x_2}  = \frac{\partial}{\partial x_2} g^{-1}(a_0 + a_1 x_1) = 0$$
Thus, $x_2$ has no effect on the mean response.

**(b) Categorical $x_2$: effect defined by contrasts**

This is the setting relevant to our simulations. Let $L_2$ denote the levels of $x_2$. For any two levels $u, \nu \in L_2$, the contrast is:
$$\Delta_{u,\upsilon}\mu(x_1, x_2) = \mu(x_1, u) - \mu(x_1, \upsilon) = g^{-1}(a_0 + a_1 x_1) - g^{-1}(a_0 + a_1 x_1) = 0.$$

#### Conclusion (main effects)

If $a_{12} = 0$ and $a_2 = 0$, then $x_2$ has no effect on the mean of the responses. Thus, whether Type I error is defined in terms of the coefficient $a_2$ or in terms of the equality of observed means across levels of $x_2$ makes no difference.

This conclusion **is not limited to means**. If the distribution of $Y | (x_1, x_2)$ depends on $x_1$, $x_2$ only through the predictor $\eta(x_1, x_2)$ --- that is, if scale or shape parameters do not depend on the factors --- then no effect of $x_2$ can appear on the mean, the median, or any other quantile of the distribution whenever $a_{12}=0$ and $a_2=0$.

### Interactions
Similarly, we show that there is no ambiguity in the interpretation of interaction effects when either $a_1 = 0$ or $a_2 =0$. Without loss of generality, assume $a_2 = 0$ and $a_1 \neq 0$. When $a_{12}=0$, 
$$
\begin{aligned}
\eta(x_1, x_2) = a_0 + a_1 x_1, \qquad
\mu(x_1, x_2) = g^{-1}(a_0 + a_1 x_1)
\end{aligned}
$$

A common definition of interaction on the response scale [@Ai:2003] is that the marginal effect of $x_1$ depends on $x_2$. We examine two situations:

**(a) Continuous predictors: interaction defined by a cross-partial derivative.**

Assuming $g^{-1}$ is twice differentiable, an interaction is present if $\frac{\partial^2 \mu}{\partial x_2 \partial x_1} \neq 0$.

We compute:
$$\frac{\partial \mu}{\partial x_1} = \dot{g}^{-1}(\eta) \cdot a_1$$

and

$$\frac{\partial^2 \mu}{\partial x_2 \partial x_1} = \ddot{g}^{-1}(\eta) \cdot \frac{\partial \eta}{\partial x_2} \cdot a_1$$

But for our settings ($a_2=a_{12}=0$)

$$\frac{\partial \eta}{\partial x_2} = a_2 + a_{12}x_1 = 0$$

and therefore

$$\frac{\partial^2 \mu}{\partial x_2 \partial x_1} = 0$$

 Thus the marginal effect of $x_1$ does not vary with $x_2$, and there is no interaction.

**(b) Categorical predictors: interaction defined by double discrete differences**

Let $L_1$ and $L_2$ be the levels of $x_1$ and $x_2$, respectively. The categorical analogue of a cross-partial derivative is the double difference:

$$\Delta_{u_1, \nu_1} \Delta_{u_2, \nu_2}\mu(x_1, x_2) = [\mu(u_1,u_2) - \mu(u_1, \nu_2)] - [\mu(\nu_1,u_2) - \mu(\nu_1, \nu_2)]$$

But when $a_2 = a_{12} = 0$, $\mu(u_1,u_2) = \mu(u_1,\nu_2)$ and $\mu(\nu_1,u_2) = \mu(\nu_1,\nu_2)$, so the double difference is always zero for any levels $u_1, \nu_1 \in L_1$ and $u_2, \nu_2 \in L_2$.

#### Conclusion (interaction effects)

If either $a_1 = 0$ or $a_2 = 0$ and $a_{12} = 0$, then no interaction can be observed on the response scale. Thus Type I error definitions based on the coefficient $a_{12}$ or on observed interaction contrasts are equivalent in these settings.

>>>>

---
title: "The illusory promise of the Aligned Rank Transform"
subtitle: "Appendix I. Additional experimental results"
author: 
  - name: Theophanis Tsandilas
    orcid: 0000-0002-0158-228X
    email: theophanis.tsandilas@inria.fr
    affiliations:
      - name: Université Paris-Saclay, CNRS, Inria, LISN
        country: France
  - name: Géry Casiez
    orcid: 0000-0003-1905-815X
    email: gery.casiezuniv-lille.fr
    affiliations:
      - name: Univ. Lille, CNRS, Inria, Centrale Lille, UMR 9189 CRIStAL
        country: France
bibliography: bibliography.bib
csl: canadian-journal-of-philosophy.csl

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Useful libraries
library(crosstalk)
library(kableExtra)
library(gridExtra)
library(lmerTest)
library(tidyverse)
library(plotly)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Source code for reading data from experimental results and for plotting
source("dataReaders.R")
source("plotting.R")

library(dplyr)
library(tidyverse)
source("plotlying.R")
source("interactions_plot.R")
```

```{=html}
<style>
.math.inline .MathJax  {
  font-size: 105% !important;
}
</style>
```

We present complementary results and new experiments that investigate additional scenarios. We also compare INT and RNK with other nonparametric methods. Unless explicitly mentioned in each section, we follow the experimental methodology presented in the main article. At the end of each section, we summarize our conclusions.

## Results for $\alpha = .01$ {#alpha}
Although we only presented results for $\alpha = .05$ in the main article, we observe the same trends for other significance levels. The following figures show the Type I error rates of the methods for the $4 \times 3$ within-subjects design when $\alpha = .01$. Note that error rates are not proportional to the $\alpha$ level.
For results from other experiments, we refer readers to our raw result files.

**Main effects**. @fig-ratio-main presents Type I error rates for the main effect of $X_2$ as the magnitude of the main effect of $X_1$ increases.

::: {#fig-ratio-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix1 <- "1_test_4x3_Ratio"
alpha <- .01
distributions1 <- c("norm", "lnorm", "exp", "poisson", "binom")
dnames1 <- c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial")
df1 <- readlyData(prefix1, alpha, 1, distributions1, dnames1)

prefix2 <- "2_test_4x3_Ordinal"
distributions2 <- c("likert5B")
dnames2 <- c("Ordinal (5 levels)")
df2 <- readlyData(prefix2, alpha, 1, distributions2, dnames2)

df <- rbind(df1,df2)

plotlyError(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 27.1)
```
Type I error rates ($\alpha = .01$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$
:::

**Interaction effects**. @fig-ratio-interaction-1 presents Type I error rates for the interaction effect $X_1 \times X_2$, when the main effect on $X_2$ is zero while the main effect on $X_1$ increases.

::: {#fig-ratio-interaction-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotlyError(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 27.1)
```
Type I error rates ($\alpha = .01$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ 
:::

**Interactions under parallel main effects**. We also present results for interaction effects when the two main effects change in parallel. As we explain in the main paper, these results require special attention, as error rates largely depend on the way we define the null hypothesis for interactions.

::: {#fig-ratio-interaction-2}
```{r, echo=FALSE, message=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df1 <- readlyData(prefix1, alpha, 0, distributions1, dnames1)
df2 <- readlyData(prefix2, alpha, 0, distributions2, dnames2)
df <- rbind(df1,df2)

plotlyError(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 104)
```
Type I error rates ($\alpha = .01$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$. *Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$. A different definition may lead to different results.* 
:::

### Conclusion
Results for $\alpha = .01$ are consistent with our results for $\alpha = .05$, confirming that ART is not a robust method.

## Missing data {#missing}
We evaluate how missing data can affect the performance of the four methods. Specifically, we study a scenario, where a random sample of $10\%$ of the observations is missing. Missing observations lead to unbalanced designs. However, we emphasize that our scenario does not cover systematic imbalances due to missing data for specific levels of a factor.

**Main effects**. @fig-missing-main presents Type I error rates for the main effect of $X_2$ as the magnitude of the main effect of $X_1$ increases. We observe that missing data make the error rate for ART to increase even further, with a larger increase for the mixed design. This is also the case for the normal distribution. In contrast, the accuracy of the three other methods does not seem to be affected.

::: {#fig-missing-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Appendix_test_missing"
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

df <- readData(prefix, n = 20, alpha = .05, effectType = 1, distributions)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 98)

```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
:::

**Interaction effects**. We also present Type I error rates for the interaction effect in the presence of a single main effect or two parallel main effects. The error levels for all methods, including ART, are now very similar to the ones observed with no missing data.

::: {#fig-missing-interaction1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 98)

```
Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
:::

### Conclusion
ART is sensitive to the presence of missing data when at least $10\%$ of the observations are missing. Specifically, we found that its Type I error rate for main effects increases further, even with normal distributions. The other methods do not seem to be affected. However, we emphasize that we only tested missing data that are randomly drawn from the complete set of observations. Results might be different if there were systematic bias in the data imbalance structure.

## Log-normal distributions {#log-normal}
In a different experiment, we evaluate log-normal distributions with a wider range of $\sigma$ parameters (see @fig-lognormal-distributions), in particular distributions with less variance, which exhibit a lower degree of skew.

::: {#fig-lognormal-distributions}
```{r, echo=FALSE, fig.height=1.3, fig.width = 9, warning=FALSE,  message=FALSE}
source("effects_plot.R")
plotEffectsLognormal(nlevels = 2, effectSize = 2)
```
Log-normal distributions that we evaluate, shown here for a factor with two levels and a magnitude of effect $a_{i} = 2$. 
:::

**Main effects**. @fig-lognormal-main presents our results on Type I error rates for main effects. As expected, ART's inflation of error rates is less serious when distributions are closer to normal, while the problem becomes worse as distributions are more skewed.

::: {#fig-lognormal-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Appendix_test_lognormal"
distributions = c("lnorm-0.2", "lnorm-0.4", "lnorm-0.6", "lnorm-0.8", "lnorm-1.0", "lnorm-1.2")
dnames = c("sd = 0.2", "sd = 0.4", "sd = 0.6", "sd = 0.8", "sd = 1.0", "sd = 1.2")

df <- readData(prefix, n = 20, alpha = .05, effectType = 1, distributions)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 69)

```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
:::

**Interaction effects**. We observe similar patterns for the Type I error rate of the interaction effect in the presence of a single main effect.

::: {#fig-lognormal-interaction1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 69)

```
Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
:::

**Interactions under parallel main effects**. We also present results when the two main effects increase in parallel. These results require again careful interpretation, as the null hypothesis of interest may be different for each method. It is important to note that even slight departures from normality (e.g., when $\sigma = 0.2$) make the interpretation of interactions problematic.

::: {#fig-lognormal-interaction2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readData(prefix, n = 20, alpha = .05, effectType = 0, distributions)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 103)

```
Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitudes $a_{1}$ and $a_{2}$ of the main effects ($n=20$). *Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$. A different definition may lead to different results.* 
:::

### Conclusion
ART's robustness issues become less severe as log-normal distributions become less skewed and approach normality. However, interpretation issues for interactions (in the presence of parallel main effects) can arise even when distributions are only slightly skewed.

## Log-normal distributions with equal variances
As we explain in the article, skewed distributions naturally lead to unequal variances, with the standard deviation of response times typically proportional to the mean. Since ART confounds effects under conditions of unequal variance, an important question remains: *Does ART’s difficulty with continuous skewed distributions stem solely from issues of heteroscedasticity?*

To address this, we present results from an additional experiment using log-normal distributions in which we directly control the population parameters of the observed distributions (rather than those of the latent variable) to ensure constant variances. Specifically, given a mean $\mu$ and a fixed standard deviation $\sigma$ for the responses, the corresponding parameters of the log-normal distribution are:

$\sigma_{log} = \sqrt{log(1 + \frac{\sigma^2}{\mu^2})}$ and $\mu_{log} = \log(\frac{\mu^2}{\sqrt{\sigma^2 + \mu^2}})$

We again consider a $4 \times 3$ within-subjects design with $n = 20$, setting the main effect of the first factor to $a_1 = 2$. We then measure the Power (effect of $X_1$) and Type I error rates (effects $X_2$ and $X_1 \times X_2$) of the methods across three standard deviations: $\sigma = 1$, $\sigma = 2$, and $\sigma = 3$.

The following table presents our results. Once again, ART inflates Type I error rates for both the effect of $X_2$ and its interaction with $X_1$, with error rates increasing as the common standard deviation grows. In contrast, PAR shows a loss in statistical power, while RNK and INT remain robust.

::: {#tbl-homoscedastic layout-ncol=3}
|     | PAR | RNK | INT | ART |
|:-----:|----:|----:|----:|----:|
|$X_1$| 100 | 100 | 100 | 100 |
|$X_2$| 5.0 | 4.9 | 5.4 | 7.6 |
|$X_1\times X_2$| 3.6 | 4.2 | 4.6 | 5.7 |

: $\sigma = 1$

| PAR | RNK | INT | ART |
|----:|----:|----:|----:|
| 98.8 | 100 | 100 | 100 |
| 4.0 | 5.1  | 5.3 | 12.1 |
| 3.5 | 5.3 |  5.1 | 9.3  |

: $\sigma = 2$

| PAR | RNK | INT | ART |
|----:|----:|----:|----:|
| 94.0 | 100 | 100 | 100 |
| 3.6 | 4.4  | 4.7 | 17.8 |
| 3.0 | 4.7  | 5.0 | 11.2 |

: $\sigma = 3$

Percentage of positives (%) over 5000 repetitions ($\alpha = .05$). For $X_1$, values reflect statistical power. For $X_2$ and $X_1 \times X_2$, they represent Type I error rates.
:::

### Conclusion
Heteroscedasticity alone cannot explain ART's failures under log-normal distributions. ART's alignment method is based on the assumption that the shape of the distribution (log-normal here) is identical across all factor levels. This is not the case in the above scenarios where distribution shapes change across the levels of $X_1$, causing the method to break down.

## Binomial distributions {#binomial}
We also evaluate a wider range of parameters for the binomial distribution. We focus on the lower range of probabilities $p$. However, we expect results to be identical for their symmetric probabilities $1-p$. Specifically, we test $p=.05$, $.1$, and $.2$, and for each, we consider $k=5$ and $10$ task repetitions (Bernoulli trials).

**Main effects**. We present our results for the main effect in @fig-binom-main. We observe that ART's Type I error rates increase as the number of repetitions decreases and the probability of success approaches zero, reaching very high levels when $k=5$ and $p=.05$. This trend is consistent across designs. The other methods maintain low error rates. However, their error rates fall below nominal levels when the magnitude of the effect on $X_1$ grows beyond a certain threshold, indicating a loss of power in these cases.

::: {#fig-binom-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Appendix_test_binom"
distributions = c("binom-5-05", "binom-5-10", "binom-5-20", "binom-10-05", "binom-10-10", "binom-10-20")
dnames = c("k=5, p=5%", "k=5, p=10%", "k=5, p=20%", "k=10, p=5%", "k=10, p=10%", "k=10, p=20%")

df <- readData(prefix, n = 20, alpha = .05, effectType = 1, distributions)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 98)

```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
:::

**Interaction effects**. @fig-binom-interaction1 shows similar patterns for the Type I error rate of the interaction effect in the presence of a single main effect.

::: {#fig-binom-interaction1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 98)

```
Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
:::

**Interactions under parallel main effects**. When both main effects exceed a certain level (see @fig-binom-interaction2), all methods fail to control the Type I error rate. Whether this inflation is due to interpretation issues or to a lack of robustness in the methods themselves, these results once again highlight that testing interactions in such scenarios is highly problematic.

::: {#fig-binom-interaction2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readData(prefix, n = 20, alpha = .05, effectType = 0, distributions)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 103)

```
Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitudes $a_{1}$ and $a_{2}$ of the main effects ($n=20$). *Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$. A different definition may lead to different results.* 
:::

### Conclusion
ART is extremely problematic under binomial distributions, raising Type I error at very high levels even when effects are null across all factors. We also observe that testing interactions in the presence of parallel main effects can be problematic for all other methods.

## Ordinal data {#ordinal}
Given the frequent use of ART with ordinal data, we evaluate our complete set of ordinal scales, based on both equidistant and flexible thresholds, with additional experimental designs.

**Main effects**. @fig-ordinal-designs-main presents Type I error rates for the main effect. ART preserves error rates at nominal levels under the $2 \times 3$ between-subjects design and the $2 \times 3$ mixed design, as long as thresholds are equidistant. Under the two within-subject designs, it inflates error rates, especially when there are fewer ordinal levels with flexible thresholds.

::: {#fig-ordinal-designs-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Appendix_test_ordinal"
alpha = 0.05

distributions <- c("likert5", "likert5B", "likert7", "likert7B", "likert11", "likert11B")
dnames <- c("5 - equidistant", "5 - flexible", "7 - equidistant", "7 - flexible", "11 - equidistant", "11 - flexible")

df <- readData(prefix, n = 20, alpha, effectType = 1, distributions)
df <- reshapeByDesign(df, dnames)

plotlyErrorByDesign(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 62)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$)
:::

**Interaction effects**. We present Type I error rates for the interaction effect in the presence of a single main effect. These results lead to similar conclusions. Even in cases where ART keeps error rates close to nominal levels (e.g., under the between-subjects design with equidistance thresholds), the performance of PAR is constantly better.

::: {#fig-ordinal-designs-interaction-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotlyErrorByDesign(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 62)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$) 
:::

**Interactions under parallel main effects**. Finally, we show results for scenarios where main effects increase in parallel.

::: {#fig-ordinal-designs-interaction-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readData(prefix, n = 20, alpha, effectType = 0, distributions)
df <- reshapeByDesign(df, dnames)
plotlyErrorByDesign(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$  ($n = 20$). *Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$. A different definition may lead to different results.* 
:::

### Conclusion
ART's inflation of Type I error rates with ordinal data is confirmed across a range of designs. For the between-subjects and mixed designs, the problem primarily concerns ordinal scales with flexible thresholds. For the within-subjects designs, ART also inflates error rates for scales with equidistant thresholds, particularly when the number of levels is as low as five or seven. Again, all methods may fail to correctly infer interactions when parallel main effects exceed a certain threshold.

## Main effects in the presence of interactions {#removable-main}
In all experiments assessing Type I error rates reported in our article, we assumed no interaction effects. However, we also need to understand whether weak or strong interaction effects could affect the sensitivity of the methods in detecting main effects. Unfortunately, the interpretation of main effects may become ambiguous when interactions are present.

### Illustrative example 
To understand the problem, let us take a [dataset](removable_main.csv) from a fictional experiment (within-participants design with $n = 24$) that evaluates the performance of two techniques (*Tech A* and *Tech B*) under two task difficulty levels (*easy* vs. *hard*). @fig-example-interactions visualizes the means for each combination of the levels of the factors, using the original scale of measurements (left) and a logarithmic scale (right). Note that time measurements were drawn from log-normal distributions.

::: {#fig-example-interactions}
```{r, echo=FALSE, message=FALSE, fig.height=3.5, fig.width = 9, warning=FALSE}
df <- read.csv("removable_main.csv", sep=",", header=TRUE, strip.white=TRUE)
dftime <- aggregate(Time ~ Difficulty+Technique, data = df, mean)

fig1 <- createInteractionPlot(dftime) %>% layout( yaxis = list(hoverformat = '.2f'))
fig2 <- createInteractionPlot(dftime, logscale = TRUE) %>% layout( yaxis = list(hoverformat = '.2f'))

fig <- subplot(fig1, fig2, titleY = TRUE, titleX = TRUE, margin = 0.06) %>% layout(hovermode = 'x', dragmode='pan')

fig
```
The line charts visualize the effects of task difficulty (*easy* vs. *hard*) and technique (*Tech* A vs. *Tech B*) for task completion time. All data points represent group means. 
:::

We observe a clear main effect of *Difficulty* and a strong interaction effect, regardless of the scale. However, the main effect of *Technique* largely depends on the scale used to present the results. Can we then conclude that *Technique A* is overall faster than *Technique B*?

Below, we present the results of the analysis using different methods.

|        | PAR  | LOG  | ART | RNK | INT |
|--------|------|------|-----|-----|-----|
| Difficulty  | $7.8 \times 10^{-8}$ | $4.3 \times 10^{-16}$  |  $8.6 \times 10^{-12}$ | $2.5 \times 10^{-16}$ | $1.1 \times 10^{-15}$ |
| Technique   | $.024$ | $.81$  | $.00017$ | $.69$ | $.69$ |
| Difficulty $\times$ Technique | $7.6 \times 10^{-6}$ | $4.1 \times 10^{-9}$ | $9.4 \times 10^{-8}$ | $1.1 \times 10^{-9}$ | $1.3 \times 10^{-9}$ |
: *p*-values for main and interaction effects {.sm}

PAR detects a main effect of *Technique* ($\alpha = .05$), as does ART, which yields an even smaller $p$-value. In contrast, LOG, RNK, and INT do not find no evidence for such an effect. As with *removable interactions*, we can speak about *removable main effects* in this case.

As a general principle, it makes little sense to interpret main effects when crossing patterns emerge due to a strong interaction. In this scenario, the researchers should instead compare techniques within each level of *Difficulty*. Accordingly, we can conclude that *Technique A* is slower on *easy* tasks but faster on *hard* tasks in this experiment.

### Experiment
To better understand how the different methods detect main in the presence of interactions, we conduct an experiment.

We focus again on two-factor experimental designs, setting the sample size to $n = 20$. To simulate populations in which interactions emerge in the absence of main effects, we examine perfectly symmetric cross-interactions. To this end, we slightly change the method we use to encode the levels of each factor, such that levels are uniformly positioned around 0. For a factor with three levels, we numerically encode the levels as $\{-0.5, 0, 0.5\}$. For a factor with four levels, we encode them as $\{-0.5, -0.1667, 0.1667, 0.5\}$.

Since the interpretation of main effects can be ambiguous, we define Type I errors with respect to the coefficients of the linear model. Accordingly, the null hypothesis is either $a_1 = 0$ or $a_2 = 0$, depending on the variable of interest.

**Interaction effect only**. We first examine how the interaction effect alone influences the Type I error rate for $X_2$. The results are shown in @fig-interactions-main-1. We emphasize that interpretation issues arise only in the case of the $4 \times 3$ design. In the other two designs, $X_1$ has only two levels, and due to their symmetry, the observed main effects should be null --- regardless of how we define the null hypothesis.

In the $4 \times 3$ design, error rates for PAR and ART grow rapidly as the interaction effect becomes large under the log-normal and exponential distributions. This is largely due to interpretation issues, as illustrated in our example. However, ART also fails to control the error rate in other scenarios. Finally, under the binomial and ordinal scales, we observe that RNK and INT may also inflate Type I error rates when the interaction effect becomes sufficiently large.

::: {#fig-interactions-main-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
alpha <- .05

prefix <- "Appendix_test_growing-interactions"
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

df <- readData(prefix, n = 20, alpha, effectType = 0, distributions)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of interaction effect", var = "rateX1", xvar = "effectX1X2", max = 104)

```
Type I error rates ($\alpha = .05$) for the **main effect of $X_1$** as a function of the magnitude $a_{12}$ of the interaction effect $X_1 \times X_2$ ($n=20$). *Note: Type I errors are defined based on the null hypothesis $a_{1} = 0$. For the $4 \times 3$ design, a different definition may lead to different results. However, the two other designs, where $X_1$ has only two levels, Type I error rates are the same, regardless of the definition we use.* 
:::

We also examine error rates for the main effect of $X_2$. Now, interpretation issues affect the results across all three designs, leading to similar patterns.

::: {#fig-interactions-main-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotlyErrorByDesign2(df, xlab = "magnitude of interaction effect", var = "rateX2", xvar = "effectX1X2", max = 104)

```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_{12}$ of the interaction effect $X_1 \times X_2$ ($n=20$). *Note: Type I errors are defined based on the null hypothesis $a_{2} = 0$. A different definition may lead to different results.*
:::

**Interaction effect combined with main effect**. Finally, we evaluate the Type I error rate on $X_2$ when the interaction effect is combined with a main effect on $X_1$. @fig-interactions-main-3 presents our results.

The error rates of ART and PAR now increase dramatically across all non-normal distributions and all three designs --- a result that is, in part, due to interpretation issues. RNK and INT are also affected: their error rates become extremely low under continuous distributions, suggesting a lack of power to detect small main effects when strong effects of other factors are combined with strong interactions. In contrast, under the binomial and ordinal scales, both methods exhibit inflated Type I error rates. Interestingly, RNK demonstrates the best overall performance in these tests.

::: {#fig-interactions-main-3}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readData(prefix, n = 20, alpha, effectType = 6, distributions)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of effects", var = "rateX2", xvar = "effectX1X2", max = 104)

```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitudes $a_1 = a_{12}$ of a combined main effect on $X_1$ and an interaction effect $X_1 \times X_2$ ($n=20$). *Note: Type I errors are defined based on the null hypothesis $a_{2} = 0$. A different definition may lead to different results.*
:::

### Conclusion
When strong interactions are present, the interpretation of main effects can become ambiguous, especially when the interaction is combined with a non-zero main effect. ART is more sensitive to these issues than other methods and may even inflate Type I error rates in scenarios where interpretation problems are not expected. More broadly, main effects should be interpreted with caution in the presence of strong interactions.

## ART with median alignment {#median-alignment}
We evaluate a modified implementation of ART (ART-MED), where we use medians instead of means to align ranks. This approach draws inspiration from results by @Salter:1993, showing that median alignment corrects ART's instable behavior under the Cauchy distribution. We only test the $4 \times 3$ within-participants design for sample sizes $n=10$, $20$, and $30$. For this experiment, we omit the RNK method and only present results for non-normal distributions.

We emphasize that @Salter:1993 only apply mean and median alignment to interactions. Our implementation for main effects is based on the alignment approach of @wobbrock:2011, where we simply replace means by medians --- we are not aware of more adapted methods.

**Main effects**. Our results presented in @fig-median-main demonstrate that median alignment (ART-MED) --- or at least our implementation of the method --- is not appropriate for testing main effects. Although Type I error rates are now lower for the Cauchy distribution compared to the original method, they are still above nominal levels. In addition, they are significantly higher for all other distributions.

::: {#fig-median-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Appendix_test_4x3_ART-Median"
alpha <- .05

methods = c("PAR", "INT", "ART", "ART-MED")
palette = c("#888888", "#009E73", "#FF5E00", "#FFA27F")
symbols = c("asterisk",  "star-diamond", "star-triangle-up", "star-triangle-down")

distributions <- c("lnorm", "exp", "cauchy", "poisson", "binom", "likert5B")
dnames <- c("Log-normal", "Exponential", "Cauchy", "Poisson", "Binomial", "Ordinal (5 levels)")
df <- readlyData(prefix, alpha, 1, distributions, dnames, methods = methods)
